---
title: "Project Proposal and EDA"
author: "Gunpreet Singh, Gurleen Pelia and Mariyam Ahmed"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

# **Urban Sprawl using Land Cover Data of Canada**

## **Purpose of Project**

#### Land Cover Data :

Canada's land cover data is important for several reasons. By offering
comprehensive information about ecosystems and forest cover, it plays a
crucial role in environmental management and conservation, enabling the
preservation of biodiversity and sustainable forest management. Studies
on climate change require these data in order to evaluate the effects on
ecosystems, vegetation patterns, and carbon sequestration. Land cover
data helps manage the risk of disasters, identify locations that are
ideal for development while maintaining green spaces, and promote
sustainable growth in urban planning and development. All things
considered, land cover data is critical to comprehending Canada's
landscapes, aiding in conservation efforts, and advancing sustainable
development in a number of industries.

Land cover data and urban sprawl are intricately related in thus project
that involve studying urban expansion and its impacts on the
environment. Understanding this relationship is crucial for analyzing
patterns, predicting future changes, and planning sustainable urban
development.

#### Urban Sprawl :

Urban sprawl refers to the unplanned and uncontrolled growth of urban
areas into neighboring rural areas, which leads to dispersed,
low-density development. Longer commutes, a greater reliance on cars,
and higher traffic congestion are the results of this phenomenon. It
destroys open areas, farms, and natural habitats, upsetting biodiversity
and local ecosystems. Infrastructure is also strained by urban sprawl,
necessitating large-scale, expensive expenditures in public utilities,
roads, and other services. It also adds to environmental problems
including pollution and higher emissions of greenhouse gases. Overall,
urban sprawl poses serious obstacles to environmental preservation and
sustainable growth.

## How Land Cover Data Helps Study Urban Sprawl?

1.  Identifying Urban Areas:

-   Land cover data allows you to identify and delineate urban areas
    from other land cover types.
-   By classifying satellite images, you can determine the extent of
    urbanization.

2.  Temporal Analysis:

-   By comparing land cover data from different time periods, you can
    analyze changes in land use and identify trends in urban expansion.
-   This helps in understanding the rate and pattern of urban sprawl.

3.  Change Detection:

-   Land cover change detection techniques enable you to quantify the
    conversion of non-urban areas to urban areas.
-   This information is crucial for assessing the impact of urban sprawl
    on natural resources and agricultural land.

4.  Modeling and Prediction:

-   Using historical land cover data, you can develop models (like
    CA-Markov) to predict future urban growth.
-   This helps in planning and implementing sustainable urban
    development strategies.

5.  Impact Assessment:

-   Analyzing land cover changes allows you to assess the environmental
    impacts of urban sprawl, such as deforestation, loss of
    biodiversity, and soil degradation.

## **About the dataset**

The CCRS Land Cover dataset is crucial for our project due to its high
accuracy, detailed resolution, and comprehensive national coverage. This
dataset's 30-meter spatial resolution provides precise and reliable land
cover information, essential for a wide range of environmental
applications such as climate impact studies, emergency response
planning, and wildlife habitat analysis. The dataset spans multiple
years (2010, 2015, and 2020), enabling temporal analysis of land cover
changes over time. Additionally, its integration into the North American
Land Change Monitoring System (NALCMS) underscores its credibility and
relevance for both national and regional studies. The solid and accurate
data foundation provided by the CCRS Land Cover dataset guarantees that
our project is founded on a dependable resource, improving the caliber
and significance of our research and assisting with well-informed
environmental management and policy decisions.

## **Description of Data Sources**

#### Primary Data Source:

-   Source Name: Canada Centre for Remote Sensing (CCRS) Land Cover Map
-   Description: This dataset provides detailed land cover information
    for Canada at a 30 m spatial resolution. It is produced using
    observations from the Operational Land Imager (OLI) Landsat sensor
    and covers the years 2010, 2015, and 2020. The data is essential for
    a wide range of environmental applications, including climate impact
    studies, emergency response, and wildlife habitat analysis.
-   Citations: Canada Centre for Remote Sensing. "Land Cover Map of
    Canada (2010, 2015, 2020)." Produced under the North American Land
    Change Monitoring System (NALCMS).

#### Secondary Data Source:

-   Source Name: National Census Data (2023)
-   Description: This dataset includes demographic information for
    residents of Canada, such as age distribution, household income, and
    population density. It is used to analyze the relationship between
    land cover and population demographics.
-   Citations: Government of Canada. "National Census Data, 2023."
    <https://www.census.gc.ca>

Collection of Land Cover products for Canada as produced by Natural
Resources Canada using Landsat satellite imagery. This collection of
cartographic products offers classified Land Cover of Canada at a 30
metre scale, updated on a 5 year basis.

-   Landcover of Canada 2010
-   Landcover of Canada 2015
-   Landcover of Canada 2020

## **Data Pre-Processing**

#### 1.Loading and Preparing the Raster Data:

-   High-Resolution Data for Canada: Initially, you worked with
    high-resolution raster data that covered the entire country of
    Canada. This dataset likely contained detailed land use or land
    cover information across the country. \*Cropping to Fit Ontario:
    Since the analysis was focused on Ontario, you cropped the
    Canada-wide raster dataset to fit the geographic boundaries of
    Ontario. Cropping helps in reducing the size of the data and
    focusing the analysis on the region of interest.

#### 2.Downsampling Using the Aggregate Function:

-   Aggregate Function: To manage the data size and make it more
    computationally feasible, you downsampled the raster data using the
    aggregate function. Downsampling reduces the resolution of the
    raster by combining several adjacent pixels into a single pixel,
    thereby reducing the data size while preserving the overall spatial
    patterns.
-   Process: The aggregate function groups adjacent cells and calculates
    a summary statistic (such as mean, sum, or majority) for the new,
    larger cells. This process retains the essential spatial information
    but with less detail.

#### 3.Setting the Coordinate Reference System (CRS):

-   CRS for Bounding Box of Canada: The CRS defines how the
    two-dimensional, projected map in your GIS relates to real places on
    the earth. You set the CRS for the bounding box of Canada using the
    st_transform function. This ensures that the spatial data aligns
    correctly with geographic coordinates.
-   Transform Function: The st_transform function from the sf package is
    used to transform the spatial coordinates of the dataset to a
    specified CRS. This is crucial for spatial operations like cropping
    and overlaying different spatial datasets.

#### 4.Collecting Data for Toronto Boundaries Using OpenStreetMap (OSM):

-   OpenStreetMap (OSM): OSM is a collaborative project that creates a
    free, editable map of the world. You used OSM to gather the boundary
    data for Toronto.
-   Key-Value Pair for Toronto: In OSM, geographical features are
    described using key-value pairs. You likely used a query with a
    specific key (e.g., boundary) and value (e.g., Toronto) to extract
    the boundary data for Toronto.
-   Data Retrieval: The boundary data for Toronto was retrieved and
    converted into a spatial object that could be used for further
    geospatial operations.

#### 5.Cropping and Masking the Ontario Land Use Data:

-   Cropping: The Toronto boundary data obtained from OSM was used to
    crop the downsampled Ontario raster data. Cropping confines the
    raster data to the area within the Toronto boundaries, removing data
    outside this area.
-   Masking: Masking further refines the raster data by setting values
    outside the specified mask (in this case, the Toronto boundary) to
    NA or some other value indicating absence of data. This ensures that
    the analysis is strictly limited to the area of interest.
-   Process: The combination of cropping and masking results in a
    dataset that contains detailed land use information specifically for
    the Toronto area, extracted from the larger Ontario dataset.

```{r,include = FALSE,echo = FALSE,fig.align = "center", fig.height = 4}
library(raster)
library(sf)
library(tmap) # for making the nice plot
library(osmdata)
library(tidyverse)
library(terra)
library(spDataLarge)
library(stars)
library(ggplot2)
library(dplyr)
```

## **Visualizations**

```{r,echo=FALSE,fig.align = "center", fig.height = 4,include=FALSE}
land_coverage_2010 <- "/Users/gunpreetsingh/Desktop/landcover-2010-classification.tif" |>
  rast()

# Set the CRS to NAD83(CSRS) / Canada Atlas Lambert (EPSG:3979)
crs(land_coverage_2010) <- "EPSG:3979"

# Define the extent of Ontario in lat/lon
ontario_extent_latlon <- ext(-95.1537, -74.3202, 41.6766, 56.8566)

# Create a SpatVector with the extent in lat/lon
ontario_extent_vect <- vect(ontario_extent_latlon, crs = "EPSG:4326")

# Transform the extent to the raster's CRS
ontario_extent_transformed <- project(ontario_extent_vect, "EPSG:3979")

# Extract the transformed extent coordinates
ontario_extent <- ext(ontario_extent_transformed)

# Crop the raster to the transformed extent
land_coverage_2010_cropped <- crop(land_coverage_2010, ontario_extent)

# Downsample the cropped raster (optional)
land_coverage_2010_downsampled <- aggregate(land_coverage_2010_cropped, fact = 16)
```

```{r,echo=FALSE}

color_palette <-  c("#115f9a", "#1984c5", "#22a7f0", "#48b5c4", "#76c68f", "#a6d75b", "#c9e52f", "#d0ee11", "#d0f400")

#Plotting the Downsampled ontario raster data
tm_shape(land_coverage_2010_downsampled) +
  tm_raster(palette = color_palette)+
  tm_layout(title = "Land Coverage Ontario - 2010", legend.outside = TRUE)

```

This map shows land coverage in Ontario, Canada, for the year 2010,
using a blue-to-yellow color palette to represent different coverage
ranges. Dark blue indicates areas with low coverage (0-5), while yellow
represents areas with high coverage (15-20). The detailed legend and
color scheme highlight spatial variations, with the southern part of
Ontario, around the Great Lakes, shown in more detail due to its higher
population density and diverse land use patterns. The map serves as a
useful tool for environmental analysis, urban planning, and research by
providing a clear visual representation of land cover distribution
across the region.

```{r,echo=FALSE,include = FALSE}
# Define the extent of Toronto in lat/lon
brampton_extent_latlon <- ext(-79.8587, -79.6633, 43.6319, 43.7559)

# Create a SpatVector with the extent in lat/lon
brampton_extent_vect <- vect(brampton_extent_latlon, crs = "EPSG:4326")

# Transform the extent to the raster's CRS
brampton_extent_transformed <- project(brampton_extent_vect, "EPSG:3979")

# Extract the transformed extent coordinates
brampton_extent <- ext(brampton_extent_transformed)

# Crop the ontario data to toronto data using crop and downsampling using aggregate
land_coverage_2010_toronto <- crop(land_coverage_2010_cropped, brampton_extent) |>
  aggregate(fact = 16)

```

```{r,echo=FALSE,include=FALSE}
# Define the bounding box coordinates
brampton_bbox <- c(left = -79.8587, right = -79.6633, bottom = 43.6319, top = 43.7559)

#Extracting Toronto Boundary using OpenStreetMap and creating an sf_dataframe
brampton_area <- opq(bbox=brampton_bbox) |> 
  add_osm_feature(key="boundary", value = "administrative") |> 
  add_osm_feature(key="name",value = "Brampton") |>
  add_osm_feature(key="admin_level",value = "8") |>
  osmdata_sf() |>
  (\(x) x$osm_multipolygons)() 

#Changing the CRS of the toronto_area that was extracted from the osm
brampton_area = st_transform(brampton_area, crs = 3979)
```

```{r, echo=FALSE}
#Cropping and masking the toronto area with toronto land use raster data
brampton_area_2010_masked = crop(land_coverage_2010_toronto,brampton_area) |>
  mask(brampton_area)

#Plotting Toronto Land Use Data using tmap 
tm_shape(brampton_area_2010_masked) +
  tm_raster(palette = color_palette) +
  tm_shape(brampton_area) +
  tm_borders(col = "blue", lwd = 2) +
  tm_layout(title = "Land Coverage of Brampton 2010",legend.outside = TRUE)
```

This map displays land coverage in Toronto, Canada, for the year 2010,
using a blue-to-yellow color gradient. Dark blue represents areas with
lower land coverage values (4 to 6), while light yellow indicates higher
coverage values (16 to 18). The map highlights spatial variations across
Toronto, with detailed insights into specific regions. Blue areas likely
correspond to urban spaces or water bodies, whereas yellow areas
indicate dense vegetation or green spaces. This visualization aids urban
planners, environmental analysts, and researchers in understanding land
use patterns for informed decision-making.

## 2015

```{r,fig.align = "center", fig.height = 4, echo=FALSE,include = FALSE}
#Loading 2015 land use raster data
land_coverage_2015 <- "/Users/gunpreetsingh/Desktop/landcover-2015-classification.tif" |>
  rast()

# Set the CRS to NAD83(CSRS) / Canada Atlas Lambert (EPSG:3979)
crs(land_coverage_2015) <- "EPSG:3979"

# Crop the raster to the transformed extent
land_coverage_2015_cropped <- crop(land_coverage_2015, ontario_extent) |>
  aggregate(fact = 16)
```

```{r,echo=FALSE}
#setting a new palette for 2015 raster data
color_palette <-  c("#115f9a", "#1984c5", "#22a7f0", "#48b5c4", "#76c68f", "#a6d75b", "#c9e52f", "#d0ee11", "#d0f400")

#Plotting the 2015 land coverage data for ontario 
tm_shape(land_coverage_2015_cropped) +
  tm_raster(palette = color_palette) +
  tm_shape(brampton_area) +
  tm_borders(col = "blue", lwd = 2) +
  tm_layout(title = "Land Coverage Ontario-2015", legend.outside = TRUE)
```

This map illustrates land coverage in Ontario, Canada, for the year
2015, using a color palette ranging from red to green. The legend
indicates that red represents areas with the lowest land coverage values
(0 to 5), green represents moderate coverage values (5 to 10 and 10 to
15), and blue-green represents the highest coverage values (15 to 20).
The map provides detailed spatial variations across Ontario,
highlighting regions with different land cover densities. Red areas
likely correspond to urban or less vegetated regions, while blue-green
areas indicate dense vegetation or green spaces. This visualization aids
urban planners, environmental analysts, and researchers in understanding
land use patterns and making informed decisions.

```{r, include = FALSE,echo=FALSE}
# Crop the raster to the transformed extent
land_coverage_2015_toronto <- crop(land_coverage_2015_cropped, brampton_extent) 
```

```{r,fig.align = "center", fig.height = 4,echo=FALSE}
#Cropping and masking the toronto area with toronto land use raster data
brampton_area_2015_masked = crop(land_coverage_2015_toronto,brampton_area) |>
  mask(brampton_area)

#Plotting land use data for toronto 2015
tm_shape(brampton_area_2015_masked) +
  tm_raster(palette = color_palette) +
  tm_shape(brampton_area) +
  tm_borders(col = "blue", lwd = 2) +
  tm_layout(title = "Land Coverage of Brampton 2015",legend.outside = TRUE)
```

This map depicts land coverage in Toronto, Canada, for the year 2015,
using a varied color palette to represent different ranges of land
coverage values. The legend indicates:

Red: 4 to 6 Light blue: 6 to 8 Purple: 8 to 10 Orange: 10 to 12
Lavender: 12 to 14 Green: 14 to 16 Teal: 16 to 18 The color scheme helps
to identify spatial variations in land coverage across Toronto. Lower
values, shown in red and light blue, likely correspond to urban areas or
less vegetated spaces. Higher values, depicted in green and teal,
indicate areas with dense vegetation or green spaces. This map is useful
for urban planners, environmental analysts, and researchers, providing a
visual tool for understanding land use patterns and making informed
decisions.

## 2020

```{r,fig.align = "center", fig.height = 4,echo=FALSE,include = FALSE}
#Loading 2020 land use raster data
land_coverage_2020 <- "/Users/gunpreetsingh/Desktop/landcover-2020-classification.tif" |>
  rast()

# Set the CRS to NAD83(CSRS) / Canada Atlas Lambert (EPSG:3979)
crs(land_coverage_2020) <- "EPSG:3979"

# Crop the raster to the transformed extent
land_coverage_2020_cropped <- crop(land_coverage_2020, ontario_extent) |>
  aggregate(fact = 16)
```

```{r,echo=FALSE}
#Plotting 2020 land use Ontario data that was extracted from the previous step
color_palette_2020 <- c("#e27c7c", "#a86464", "#6d4b4b", "#503f3f", "#333333", "#3c4e4b", "#466964", "#599e94", "#6cd4c5")

#Plotting the 2020 land coverage data for ontario  
tm_shape(land_coverage_2020_cropped) +
  tm_raster(palette = color_palette_2020) +
  tm_shape(toronto_area) +
  tm_borders(col = "blue", lwd = 2) +
  tm_layout(title = "Land Coverage Ontario-2020",legend.outside = TRUE)
```

This map presents land coverage in Ontario, Canada, for the year 2020,
using a color scheme that ranges from red to teal. The legend specifies
the land coverage values associated with each color:

Red (0 to 5) Dark gray (5 to 10) Dark teal (10 to 15) Light teal (15 to
20) The map highlights spatial variations across Ontario, with red areas
representing regions with the lowest land coverage values, likely
corresponding to urban areas or regions with less vegetation. The teal
areas denote regions with the highest land coverage values, indicative
of dense vegetation or green spaces. This visualization is valuable for
urban planners, environmental analysts, and researchers, as it provides
a clear visual representation of land use patterns and aids in informed
decision-making related to environmental management and urban
development.

```{r, echo = FALSE,include = FALSE,echo=FALSE}
# Crop the raster to the transformed extent
land_coverage_2020_toronto <- crop(land_coverage_2020_cropped, toronto_extent) 
```

```{r,fig.align = "center", fig.height = 4,echo=FALSE}
#Cropping and masking the toronto area with toronto land use raster data
toronto_area_2020_masked = crop(land_coverage_2020_toronto,toronto_area) |>
  mask(toronto_area)

#Plotting land use data for toronto 2020
tm_shape(toronto_area_2020_masked) +
  tm_raster(palette = color_palette_2020) +
  tm_shape(toronto_area) +
  tm_borders(col = "blue", lwd = 2) +
  tm_layout(title = "Land Coverage of Toronto 2020",legend.outside = TRUE)
```

This map displays land coverage in Toronto, Canada, for the year 2020,
using a color scheme from red to teal. Red areas represent the lowest
land coverage values (4 to 6), various shades of gray indicate moderate
values (6 to 12), and teal areas show the highest values (12 to 18). The
map highlights spatial variations in land coverage across Toronto, with
lower values likely corresponding to urban areas and higher values
indicating dense vegetation or green spaces. This visualization aids
urban planners, environmental analysts, and researchers in understanding
land use patterns for informed decision-making.

## **Analysis Methods/Modelling**

We are planning to use A-Markov modeling by taking reference from 6
different research paper which are mentioned below. This model is a
powerful method for simulating and predicting land use changes over time
by combining the strengths of Cellular Automata (CA) and Markov Chains.

#### Cellular Automata (CA) :

Cellular Automata are dynamic models that simulate the evolution of
spatial patterns over time. They work on a grid of cells, where each
cell can be in one of a finite number of states (e.g., different land
cover types). The state of each cell at the next time step is determined
by a set of rules that consider the current state of the cell and the
states of its neighboring cells.

#### Mathematical Formulation:

-   **State Space:** $S = \{s_1, s_2, ..., s_n\}$

    Each cell $c$ can be in one of the states $s_i$ at time $t$.

-   **Neighborhood:** The neighborhood $N(c)$ of cell $c$ includes the
    cells that influence the state of $c$.

-   **Transition Rules:** The transition rules $T$ determine the state
    of a cell based on the states of its neighbors: $$
    s_c(t+1) = T(s_c(t), N(s_c(t)))
    $$

#### Markov Chains:

Markov Chains are stochastic models that describe the probability of
transitioning from one state to another over time. They are used to
model the likelihood of land cover change based on historical data. The
transition probabilities are typically derived from observed changes in
land cover over a specified period.

#### Mathematical Formulation:

-   **State Vector:** $\mathbf{p}(t)$

    A vector representing the probability distribution of the states at
    time $t$.

-   **Transition Matrix:** $\mathbf{P}$

    A matrix where $P_{ij}$ represents the probability of transitioning
    from state $i$ to state $j$: $$
    \mathbf{P} =
    \begin{pmatrix}
    P_{11} & P_{12} & \cdots & P_{1n} \\
    P_{21} & P_{22} & \cdots & P_{2n} \\
    \vdots & \vdots & \ddots & \vdots \\
    P_{n1} & P_{n2} & \cdots & P_{nn}
    \end{pmatrix}
    $$

-   **Next State Probability:** The state vector at time $t+1$ is
    obtained by multiplying the state vector at time $t$ with the
    transition matrix: $$
    \mathbf{p}(t+1) = \mathbf{p}(t) \mathbf{P}
    $$

#### CA-Markov Model:

The CA-Markov model integrates Cellular Automata and Markov Chains to
predict future land cover changes. The Markov Chain component provides
the transition probabilities, while the Cellular Automata component
spatially allocates these changes based on local neighborhood rules.

#### Mathematical Formulation:

1.  **Initial State:** Start with an initial land cover map $S_0$.

2.  **Transition Matrix:** Compute the transition matrix $\mathbf{P}$
    from historical land cover maps.

3.  **Neighborhood Influence:** Apply the CA rules to incorporate
    spatial dependencies.

The state of a cell at time $t+1$ depends on:

-   The transition probabilities from the Markov Chain.
-   The neighborhood configuration from the CA rules.

#### Steps to Implement CA-Markov Modeling

1.  Data Preparation:

-   Collect temporal and spatial land cover data
-   Preprocess the data

2.  Land Cover Classification:

-   Classify the land cover data into different categories

3.  Transition Probability Matrix:

-   Calculate the transition probabilities using historical land cover
    data.

4.  Define CA Rules:

-   Establishe rules that determine how cells change state based on
    their neighbors.

5.  Model Calibration and Validation:

-   Validate the model using a known time period to ensure accuracy.
-   Adjust parameters to improve the model’s predictive capability.

6.  Prediction:

-   Use the calibrated model to predict future land cover changes

## **References**

1.  Shikhar Deep, Akansha Saklani. (2014). "Urban sprawl modeling using
    cellular automata." The Egyptian Journal of Remote Sensing and Space
    Sciences, 17, 179–187. DOI: 10.1016/j.ejrs.2014.07.001.

2.  Mussie G. Tewolde, Pedro Cabral. (2011). "Urban Sprawl Analysis and
    Modeling in Asmara, Eritrea." Remote Sens., 3, 2148-2165. DOI:
    10.3390/rs3102148.

3.  Parvinnezhad, Davoud, et al. (2021). "A modified spatial entropy for
    urban sprawl assessment." Geocarto International, 36(16), 1804-1819.

4.  Zeng, Chen, et al. (2015). "Characterization and spatial modeling of
    urban sprawl in the Wuhan Metropolitan Area, China." International
    Journal of Applied Earth Observation and Geoinformation, 34, 10-24.

5.  Chen, Danling, et al. (2021). "How urban sprawl influences
    eco-environmental quality: Empirical research in China by using the
    Spatial Durbin model." Ecological Indicators, 131, 108113.

6.  Baqa, Muhammad Fahad, et al. (2021). "Monitoring and modeling the
    patterns and trends of urban growth using urban sprawl matrix and
    CA-Markov model: A case study of Karachi, Pakistan." Land, 10(7),

    700. 
